Best Practices in Latent Class Analysis using Free Open Source Software

### 2023-01-23 manuscript.tex
edits by Mauricio

Latent class analysis (LCA) is an umbrella term that refers to a number
of techniques for estimating unobserved group membership based on a
parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile
analysis. Vermunt et al. (2004) defined LCA more generally as
virtually any statistical model where ``some of the parameters {[}\ldots{]}
differ across unobserved subgroups''.

Despite the popularity of LCA,
two challenges impede broader adoption and correct application of the method.
The first challenge is that most existing software that implements these models is commercial and closed-source.
The second challenge is that there is a lack of standards for estimating and reporting LCA, and some myths and misunderstandings have been perpetuated in prior literature.
This introduces a risk of misapplications of the technique,
and complicates manuscript review and assessment of the quality of published studies.
The present paper seeks to address these challenges by introducing updated guidelines for estimatin and reporting LCA,
based on current best practice.
Furthermore, meeting the increasing demand for open source research software,
this paper introduces new functionality in the R-package \texttt{tidySEM} (\textbf{vanlissaTidySEMTidyStructural2022?}) to perform LCA using free open source software (FOSS).

@@

There are several compelling reasons why researchers should prefer free open source software (FOSS) instead.
FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
This is invaluable for reducing gatekeeping of researchers with limited budgets, such as those from developing countries.
Furthermore, FOSS promotes collaboration within the scientific community,
as its users can contribute to the development, maintenance, and support of the software.
Finally, FOSS promotes tranparency and reproducibility: Because the source code for FOSS is accessible,
researchers can review and verify the methods and algorithms used,
which helps to ensure the integrity and reliability of research findings.
Overall, FOSS has the potential to significantly improve the efficiency and effectiveness of scientific research, while also promoting transparency, collaboration, and inclusivity within the scientific community.

Notable FOSS for LCA includes Mclust {[}REF{]}, depmixS4 {[}REF{]}, and OpenMx (Neale et al., 2016).
A crucial limitation of these existing packages is that some have limited functionality (e.g., Mclust),
or implement specific innovations (e.g., depmixS4).
OpenMx is unique in that, like Mplus, it implement LCA in the context of a fully featured structural equation modeling framework.
Remaining obstacles to the use of \texttt{OpenMx} is that it is not very user-friendly and its functionality for LCA is poorly documented.

@@

When the focus is on the model parameters in each group, LCA can be
thought of as similar to a multi-group structural equation model. The
main distinction is that group membership is not known a-priori, but is
instead estimated -- with measurement error -- based on the data.
Whereas in a multi-group model, the data are split by group and treated
as independent samples, in LCA, all cases contribute to the estimation
of all parameters in all groups. The relative contribution of each case
to the parameters of each group is determined by that case's posterior
probability of belonging to that group.

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm that corrects for (random) measurement error.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.
These posterior class probabilities can be used to weight follow-up analyses,
or individuals can be assigned to classes based on their highest class probability.
If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
This perspective on mixture modeling is sometimes referred to as ``model-based
clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \& Raftery, 2016).
Many clustering algorithms apply some recursive splitting algorithm to the data. By
contrast ``model-based'' clustering refers to the fact that LCA estimates
cluster membership based on a parametric model.
Since a parametric model estimates only a fixed number of parameters,
it can be a relatively parsimonious solution compared to non-parametric techniques.

Finally, in the context of machine learning, LCA can be considered as an
\emph{unsupervised classification} method (Figueiredo \& Jain, 2002).
The term \emph{unsupervised} refers to the fact that the outcome variable --
true class membership -- is not known, and the term \emph{classification}
refers to the fact that the algorithm is predicting a categorical
outcome: class membership.

@@

The term LCA was initially conceived to refer to analyses with categorical,
usually binary indicators (Collins \& Lanza, 2009).
It was later generalized to refer to any technique that estimates a categorical latent variable based on a
parametric model of observed indicators, as it only refers
to the purpose of the analysis, and does not imply restrictions to the
model used, or the level of measurement of the indicators (Vermunt et al., 2004).
To prevent ambiguity, we refer to LCA with binary or ordinal indicators as \emph{LCA with ordinal indicators}.
Given the abundance of redundant terms in use for closely related types of LCA models,
we here provide a rudimentary taxonomy.

A common type of LCA is the \emph{finite Gaussian mixture model}; a
univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal
(Gaussian) distributions. The parameters of a finite Gaussian mixture
model are the means and variances of these underlying normal
distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be estimated
freely, or set to be constrained across classes (Collins \& Lanza, 2009).

@@

@@

Although LCA is often discussed as an exploratory analysis technique,
it can also be used in a confirmatory manner.
Given that LCA is similar to confirmatory factor analysis,
it can be used to similar ends:
To assess whether a theoretical measurement model holds.
This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
For example, \emph{identity status theory} posits that, at any given point in time, adolescents reside in one of four identity statuses (Marcia, 1966).
LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) (\textbf{luyckxDevelopmentalTypologiesIdentity2008?}).
Similarly, \emph{personality type theory} states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient type, which could also be restored using LCA (\textbf{donnellanResilientOvercontrolledUndercontrolled2010?}).
One challenge is that, unlike confirmatory factor analysis, LCA does not provide absolute fit indices (see Model Fit Indices).
Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
either qualitatively or quantitatively, using significance tests for pre-specified values (see Inference).
A recent study by Maene, D'hondt, Van Lissa, Thijs, and Stevens (2022) applied this principle to
test the theory that identities of Belgian adolescents with migration background could be summarized as those relating to their heritage, national and regional identity.

@@

LCA can also be used as a measurement model, similar to confirmatory factor analysis.
This is useful when class membership is measured imperfectly by several indicators:
LCA can partial out measurement error and restore most likely class membership.
For example, a recent study found discrepancies between self-reported employment status and official register data (Pavlopoulos \& Vermunt, 2015).
While it might be tempting to assume that register data is free of error, this assumption was shown to be false.
A mixture model was used to estimate ``error corrected employment status'', using self-report and registry data as indicators.
Among the outcomes of interest were random and systematic measurement error in both indicators,
which reveal strengths and weaknesses of both types of assessment.

@@

LCA can also be used to reduce the dimensionality of data
by reducing many variables to a few prototypes.
Factor analysis is also sometimes used for dimension reduction,
but only accounts for linear covariance between indicators,
whereas LCA can accommodate complex - but discrete - patterns.
For example, imagine that there exist two indicators - \(x\) and \(y\), and some individuals score high on both, some score low on both, and some score high on \(x\) but low on \(y\).
Factor analysis would only be able to capture the differences between the first two groups well; LCA captures differences between all three.
Note that this is a pragmatic use of LCA;
there is no theory about the existence of a categorical latent variable (cf.~Testing theory).
Consequently, the class solution should be treated as descriptive of the sample,
and not be ``reified'' - treated as evidence of the existence of a categorical latent variable.
The use of LCA for dimension reduction is common in longitudinal research,
where one developmental process is summarized using latent class growth analysis (LCGA),
and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy and found three groups: high, average, and low empathy (Van Lissa et al., 2015).
These groups were then used as moderator of a latent growth analysis of adolescent- and parent-reported conflict.
Results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than the other groups,
and that the low empathy group had more conflict with parents than the other groups.
Note that both these key findings are non-linear;
modeling them in a single-group model would be difficult.
LCGA greatly simplifies model specification and interpretation,
albeit at a cost of some loss of information and, potentially, generalizability.

@@

\hypertarget{class-enumeration}{%
\subsubsection{Class enumeration}\label{class-enumeration}}

Another use case is to explore heterogeneity in a population,
and determine whether the population is comprised of latent subpopulations, and if so, how many.
The primary focus in this use case is on class enumeration.
For example,
one study explored heterogeneity in the population of military service members and veterans with prior suicide attempts,
and identified two latent profiles characterized by high versus low suicide risk based on self-reported suicidal ideation and information about prior attempted methods and severity
(Gromatsky et al., 2022).

@@

Finally, LCA can be used to deal with data which violate certain assumptions.
For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) (\textbf{baughmanMixtureModelFramework2007?}).
These models estimate one class to explain the ``true zeros'',
and another class with a different distribution to explain all oter values, which can include zero.
Note that this LCA has different models for both classes,
and thus falls beyond the scope of this paper.
For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality (Alkema et al., 2022).
A mixture of two normal distributions with distinct variances described the observed distribution better.

@@

LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
Numeric variables should have an interval measurement level,
such that it is sensible to estimate means and (co)variances.
For the remaining three measurement levels,
some preprocessing is required.
The exact data types accepted by \texttt{tidySEM} at the time of writing are \texttt{numeric} and \texttt{ordered}.
To correctly specify ordered variables,
such as Likert scales
use the function \texttt{OpenMx::mxFactor()}.
Binary variables,
such as presence or absence of some symptom, are also treated as ordinal, with one category coded as zero and the other as one.
Nominal variables are converted to binary indicators of group membership via dummy coding, for example, using \texttt{mx\_dummies()}.

@@

A third consideration in preprocessing relates to the univariate distributions of indicators.
For example, the \texttt{plas\_depression} data included with \texttt{tidySEM} show extreme right skew.
These data assess depressive symptoms;
many respondents report (near) zero scores,
and some respondents report elevated scores.
To model skewed variables like this,
it makes sense to preprocess the data by transforming the raw scores to reduce skew.
Alternatively (or additionally), one can specify a model with free variances across classes:
this might result in one class with a near-zero mean and small variance to account for the fact that many respondents have no depressive symptoms,
and one class with an elevated mean and large variance to account for all respondents with non-zero symptoms.

@@

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
(Van De Schoot et al., 2017).
While we concur that investigating missingness is due diligence,
it is important to emphasize that missingness is adequately handled by default in \texttt{OpenMx}, which is the computational backend of \texttt{tidySEM}.
Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
information without imputing missing values (\textbf{endersRelativePerformanceFull2001?}).
FIML is a best-practice solution for handling missing data;
at least on par with multiple imputation (Lee \& Shi, 2021).
Multiple imputation is less suitable to LCA for two reasons. First,
because LCA methods are often computationally expensive,
and conducting them on multiple imputed datasets may be unfeasible.
Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population.

Three types of missingness have
been distinguished in the literature (Enders, 2022; Rubin, 1976): random missingness (MCAR);
missingness contingent on \emph{observed} variables that can be known and controlled for (MAR);
and missingness related to unobserved variables,
which can be neither known nor controlled for (MNAR).
A so-called ``MCAR'' test is often reported (Jamshidian \& Jalal, 2010),
but note that this name is somewhat misleading,
as a significant test indicates that missingness is MAR,
but a non-significant test statistic does not distinguish between MCAR or MNAR.
As the classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality (Little, 1988).
This assumption is tenuous in the context of LCA.
A non-parametric MCAR test may be more suitable (Jamshidian \& Jalal, 2010).
For this tutorial,
the lead author has implemented this test in the \texttt{mice} package as \texttt{mice::mcar()}.

To conclude; our recommendation is to inspect and report the proportion of missingness per variable (e.g., using \texttt{tidySEM::descriptives()}),
and to test for MAR using \texttt{mice::mcar()}.
As FIML estimation assumes that missingness is either MCAR or MAR,
one would proceed with FIML regardless of the outcome of the test.

@@

Any LCA model should be specified based on theoretical and empirical considerations.
The model parameters depend on the measurement level of the indicators.
Continuous indicators can be parametrized in terms of means, variances, and covariances.
LCGA and GMM have these same parameters,
but with respect to the latent growth variables.
Ordinal indicators can be parametrized differently;
the default approach in \texttt{tidySEM} assumes that each ordinal indicator reflects an underlying standard normal distribution, \(N(\mu = 0, \sigma = 1)\).
The parameters are ``thresholds'' that correspond to quantiles of this distribution, estimated based on the proportion of individuals in each of the
response categories of the indicator variable.
For example, a binary
indicator has a single threshold that distinguishes the two response
categories. If responses are distributed 50/50, then the corresponding
threshold would be \(t_1 = 0.00\). If the responses are distributed 60/40,
then the resulting threshold would be \(t_1 = 0.25\).
All of the aforementioned parameters can be freely estimated,
or constrained across classes, or fixed (e.g., to zero).

Prior literature emphasized the importance of considering alternative model specifications (\textbf{vandeschootGRoLTSChecklistGuidelinesReporting2017?}).
However, what has remained underemphasized is that alternative model specifications should be approached differently for exploratory versus confirmatory analyses (see Use Cases).
Most prior literature has emphasized exploratory LCA (e.g., Nylund, Asparouhov, \& Muthén, 2007).
In this context, model specification typically consists of an exhaustive grid search along several model specifications and varying numbers of classes.
A ``final'' model is then selected based on a combination of fit indices,
significance tests,
and interpretability (see Class Enumeration).

Confirmatory LCA does not require such an extensive search over the model space.
If hypothesized models have been specified a priori,
and possibly even preregistered,
it makes sense to focus on these models, rather than proceed in a purely data-driven manner.
Nonetheless, an important reason to consider alternative model specifications is that LCA lacks objective fit indices.
The fit of a theoretical model can thus only be assessed by comparison to other candidates.
For this reason, the theoretical model should still be compared to a few others.
A one-class model is a sensible benchmark,
to test whether theoretical model fits better than a model without latent classes.
Other candidates for comparison include competing theoretical models and those with a few more or less classes than the hypothesized number.
If the theoretical model has much better fit than other models,
this provides evidence for the theory.
If differences in fit are small, however, this may not do not give cause to reject a theoretical model.
The theoretical model may still be preferred, for example, to ensure comparability with prior work.
The fact that the theoretical model received little differential support should of course be discussed,
along with potential implications for future work.

A second consideration in model specification is the risk of overfitting (\textbf{hastieElementsStatisticalLearning2009?}).
Overfitting occurs when a model has many parameters relative to the number of observations.
Such a overparametrized model captures much idiosyncratic noise in a sample,
limiting its generalizability to new samples.
As explained before, the number of parameters in LCA scales with the number of estimated classes.
Consequently, LCA with many classes have a potentially very high number of parameters.
In Bayesian mixture modeling,
it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased (\textbf{rousseauAsymptoticBehaviourPosterior2011?}).
Although it is not known whether frequentist LCA exhibits the same behavior,
this suggests that low class counts might indicate overfitting.
The section on Class Enumeration addresses further ways to guard against overfitting.

@@

One problem with the random starts approach is that it is
computationally expensive to run this many replications. Moreover,
because the algorithm begins with random starting values,
it is inefficient because many marbles are likely to start far away from any valley.
Additionally, if all sets of starting values are close to a specific local optimum,
there is a risk that the global optimum is never found.
An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing (SA, \textbf{coranaMinimizingMultimodalFunctions1987?}).
SA does not behave quite like the marbles of the random starts approach;
instead, it iteratively considers some ``destination'' in the landscape,
and compares its \(-2LL\) to the current one.
If the destination \(-2LL\) is lower,
the estimator moves there.
However, if the destination \(-2LL\) is \emph{higher},
it still moves there occasionally, with a certain probability.
Think of this as occasionally flicking the marble back out of valleys,
to see if it rolls back into the current valley, or finds another,
even deeper one.
This property allows SA it to escape local optima and find the global optimum.

@@

>>>>

Latent class analysis (LCA) is an umbrella term that refers to a number
of techniques for estimating unobserved group membership based on a
parametric model of one or more observed indicators of group membership.
The types of LCA have become quite popular across scientific fields,
most notably finite Gaussian mixture modeling and latent profile
analysis. Jeroen K. Vermunt et al. (2004) defined LCA more generally as virtually
any statistical model where ``some of the parameters {[}\ldots{]} differ across
unobserved subgroups''.

Despite the popularity of LCA, two challenges impede broader adoption
and correct application of the method. The first challenge is that most
existing software that implements these models is commercial and
closed-source. The second challenge is that there is a lack of standards
for estimating and reporting LCA, and some myths and misunderstandings
have been perpetuated in prior literature. This introduces a risk of
misapplications of the technique, and complicates manuscript review and
assessment of the quality of published studies. The present paper seeks
to address these challenges by introducing updated guidelines for
estimating and reporting LCA, based on current best practice.
Furthermore, meeting the increasing demand for open source research
software, this paper introduces new functionality in the R-package
\texttt{tidySEM} (\textbf{vanlissaTidySEMTidyStructural2022?}) to perform LCA using free
open source software (FOSS).

@@

There are several compelling reasons to prefer free open source software
(FOSS) instead. FOSS is freely available for anyone to use, modify, and
distribute, which means that scientists can access and use a wide range
of powerful tools without financial impediments. This is invaluable for
reducing gatekeeping of researchers with limited budgets, such as those
from developing countries. Furthermore, FOSS promotes collaboration
within the scientific community, as its users can contribute to the
development, maintenance, and support of the software. Finally, FOSS
promotes transparency and reproducibility: Because the source code for
FOSS is accessible, researchers can review and verify the methods and
algorithms used, which helps to ensure the integrity and reliability of
research findings. Overall, FOSS has the potential to significantly
improve the efficiency and effectiveness of scientific research, while
also promoting transparency, collaboration, and inclusion within the
scientific community.

Notable FOSS for LCA includes mclust (Scrucca, Fop, Murphy, \& Raftery, 2016), depmixS4
(Visser \& Speekenbrink, 2010), and OpenMx (Neale et al., 2016). A crucial limitation of
these existing packages is that some have limited functionality (e.g.,
Mclust), or implement specific innovations (e.g., depmixS4). OpenMx is
unique in that, like Mplus, it implement LCA in the context of a fully
featured structural equation modeling framework. Remaining obstacles to
the use of \texttt{OpenMx} is that it is not very user-friendly and its
functionality for LCA is poorly documented.

@@

When the focus is on the model parameters in each group, LCA can be
thought of as similar to a multi-group structural equation model. The
main distinction is that group membership is not known a-priori, but
estimated from the data. Whereas in a multi-group model, the data are
split by group and treated as independent samples, in LCA, all cases
contribute to the estimation of all parameters in all groups. The
relative contribution of each case to the parameters of each group is
determined by that case's posterior probability of belonging to that
group.

When the focus is on each individual's estimated class membership, LCA
can be thought of as a type of clustering algorithm that corrects for
(random) measurement error. Specifically, the posterior class
probability that an individual belongs to a latent class can be computed
from the likelihood of that individual's observed data under given the
class-specific model. These posterior class probabilities can be used to
weight follow-up analyses, or individuals can be assigned to classes
based on their highest class probability. If the classes are clearly
distinct, measurement error is low, and the latter approach may be
acceptable. This perspective on mixture modeling is sometimes referred
to as ``model-based clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca et al., 2016). Many clustering algorithms apply some recursive
splitting algorithm to the data. By contrast ``model-based'' clustering
refers to the fact that LCA estimates cluster membership based on a
parametric model. Since a parametric model estimates only a fixed number
of parameters, it can be a relatively parsimonious solution compared to
non-parametric techniques.

Finally, in the context of machine learning, LCA can be considered as an
\emph{unsupervised} learning method for \emph{classification}
(Figueiredo \& Jain, 2002). The term \emph{unsupervised} refers to the
fact that the outcome variable -- true class membership -- is not known,
and the term \emph{classification} refers to the fact that the algorithm is
predicting a categorical outcome: class membership.

@@

The term LCA was initially conceived to refer to analyses with
categorical, usually binary indicators (Collins \& Lanza, 2009). It was
later generalized to refer to any technique that estimates a categorical
latent variable based on a parametric model of observed indicators, as
it only refers to the purpose of the analysis, and does not imply
restrictions to the model used, or the level of measurement of the
indicators (Jeroen K. Vermunt et al., 2004). To prevent ambiguity, we refer to LCA
with binary or ordinal indicators as \emph{LCA with ordinal indicators}.
Given the abundance of redundant terms in use for closely related types
of LCA models, we here provide a rudimentary taxonomy.

A common type of LCA is the \emph{finite Gaussian mixture model}; a
univariate analysis where the observed distribution of a single
continuous variable is assumed to result from a mixture of a known
number of normal (Gaussian) distributions. The parameters of a finite
Gaussian mixture model are the means and variances of these underlying
normal distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be
estimated freely, or set to be constrained across classes
(Collins \& Lanza, 2009).

@@

A longitudinal extension of LCA is the Hidden Markov Model (HMM, also
called Latent Transition Analysis), which estimates the change in class
membership over time with the markov assumption over time. The markov
assumption states that the class membership at time \(t\) is only affected
by the class membership at time \(t-1\). The structural part of the model
is comprise of the initial states (proportion of the sample in each
class at the first time point), and transition probabilities
(probabilities to switch class membership across time). The transition
probabilities are estimated as multinomial logistic regressions between
class memberships (Jeroen K. Vermunt, 2010).

@@

Although LCA is often discussed as an exploratory analysis technique, it
can also be used in a confirmatory manner. Given that LCA is similar to
confirmatory factor analysis, it can be used to similar ends: To assess
whether a theoretical measurement model holds. This use case is relevant
when testing a theory that postulates the existence of a categorical
latent variable. For example, \emph{identity status theory} posits that, at
any given point in time, adolescents reside in one of four identity
statuses (Marcia, 1966). LCA can be used to identify these
four statuses based on observed indicators (e.g., self-reported identity
exploration and commitment)
(\textbf{luyckxDevelopmentalTypologiesIdentity2008?}). Similarly, \emph{personality
type theory} states that dimensional differences in personality can
largely be explained by an undercontrolled, overcontrolled, and
resilient type, which could also be restored using LCA
(\textbf{donnellanResilientOvercontrolledUndercontrolled2010?}).

One challenge is that, unlike confirmatory factor analysis, LCA does not
provide absolute fit indices (see Model Fit Indices). Thus, to test a
theory, one can ascertain that the data are better described by the
number of classes dictated by theory than by a different number of
classes. Furthermore, it would be possible to test whether the observed
pattern of class-specific means corresponds to a hypothesized pattern;
either qualitatively or quantitatively, using significance tests for
pre-specified values (see Inference).

Appendix A is a tutorial to confirmatory LCA in \texttt{tidySEM}, based on a
recent study that tested the theory that identities of Belgian
adolescents with migration backgrounds could be summarized as those
relating to their heritage, national and regional identity
(Maene, D'hondt, Van Lissa, Thijs, \& Stevens, 2022).

\hypertarget{exploring-heterogeneity}{%
\subsubsection{Exploring Heterogeneity}\label{exploring-heterogeneity}}

Another use case is to explore whether a population is comprised of
latent subpopulations, and if so, how many. The primary focus in this
use case is on class enumeration. For example,
one study explored heterogeneity in the population of
military service members and veterans with prior suicide attempts, and
identified two latent profiles characterized by high versus low suicide
risk based on self-reported suicidal ideation and information about
prior attempted methods and severity (Gromatsky et al., 2022).
Knowing that the population of suicidal service members is heterogeneous
can inform future research, policy, and care provided. Appendix B offers
a tutorial for exploratory LCA, based on the aforementioned study of the
distribution of ocean microplastics (Alkema et al., 2022).

@@

LCA can also be used as a measurement model, similar to confirmatory
factor analysis. This is useful when class membership is measured
imperfectly by several indicators: LCA can partial out measurement error
and restore most likely class membership. For example, one study found
discrepancies between self-reported employment status and official
register data (Pavlopoulos \& Vermunt, 2015). While it might be tempting
to assume that register data is free of error, this assumption was shown
to be false. A mixture model was used to estimate employment status,
correcting for measurement error, from both self-report and registry
data. The study investigated random and systematic measurement error in
both indicators, which reveal strengths and weaknesses of both types of
assessment.

@@

LCA can also be used to reduce high-dimensional data to just a few
prototypes. Factor analysis is also sometimes used for dimension
reduction, but only accounts for linear covariance between indicators,
whereas LCA can accommodate complex - but discrete - patterns. For
example, imagine that there exist two indicators - \(x\) and \(y\), and some
individuals score high on both, some score low on both, and some score
high on \(x\) but low on \(y\). Factor analysis would only be able to
capture the differences between the first two groups well; LCA captures
differences between all three. Note that this is a pragmatic use of LCA;
there is no theory about the existence of a categorical latent variable
(cf.~Testing theory). Consequently, the class solution should be treated
as descriptive of the sample, and not be ``reified'' - treated as evidence
of the existence of a categorical latent variable.

The use of LCA for dimension reduction is common in longitudinal
research, where one developmental process is summarized using latent
class growth analysis (LCGA), and the resulting categories are then used
as a moderator of another developmental process in a multi-group model.
For example, one study conducted a LCGA of adolescent empathy
development on two dimensions of empathy and found three groups: high,
average, and low empathy (Van Lissa et al., 2015). These groups were
then used as moderator of a latent growth analysis of adolescent- and
parent-reported conflict. Results showed that the high-empathy group
evidenced greater adolescent-parent agreement about the incidence of
conflict than the other groups, and that the low empathy group had more
conflict with parents than the other groups. Note that both these key
findings are non-linear; modeling them in a single-group model would be
difficult. LCGA greatly simplifies model specification and
interpretation, albeit at a cost of some loss of information and,
potentially, generalizability. Appendix C is a tutorial on LCGA, based
on ongoing research on trajectories of depressive symptoms in military
service members after deployment.

@@

@@

Finally, LCA can be used to deal with data which violate certain
assumptions. For example, zero-inflated and hurdle models are special
cases of LCA for dealing data with a very high proportion of zeros
(extreme right skew) (\textbf{baughmanMixtureModelFramework2007?}). These models
estimate one class to explain the ``true zeros'', and another class with a
different distribution to explain all other values, which can include
zero. Note that this LCA has different models for both classes, and thus
falls beyond the scope of this paper.
For instance, the aforementioned study on marine plastic noted that many
prior studies had treated the distribution of particles as
(multivariate) normal, despite clear deviations from normality
(Alkema et al., 2022). A mixture of two normal distributions with
distinct variances described the observed distribution better.

@@

LCA can accommodate variables of different measurement levels:
continuous, binary, nominal, and ordinal. Numeric variables should have
an interval measurement level, such that it is sensible to estimate
means and (co)variances. For the remaining three measurement levels,
some preprocessing is required. The exact data types accepted by
\texttt{tidySEM} at the time of writing are \texttt{numeric} and \texttt{ordered}. To
correctly specify ordered variables such as Likert scales, use the
function \texttt{OpenMx::mxFactor()}. Binary variables, such as presence or
absence of some symptom, are also treated as ordinal, with one category
coded as zero and the other as one. Nominal variables are converted to
binary indicators of group membership via dummy coding, for example,
using \texttt{mx\_dummies()}.

@@

A third consideration in preprocessing relates to the univariate
distributions of indicators. For example, the \texttt{plas\_depression} data
show extreme right skew in depressive symptoms: most respondents report
few symptoms, and few respondents report many symptoms. Similarly, the
\texttt{alkema\_microplastics} data are extremely right-skewed: Very small
plastic fragments abound in the oceans, and large pieces are rare. These
distributional characteristics must be taken into account either in
preprocessing, or model specification. In preprocessing, the data can be
transformed to reduce skew. After reducing skew, a simpler LCA model may
fit the data well. Alternatively, one can specify a model that
accommodates the distributional idiosyncrasies. For example, allowing
free variances across classes might result in one class with a near-zero
mean and small variance to account for the fact that most respondents
have few depressive symptoms, and one class with an elevated mean and
large variance to account for all respondents with non-zero symptoms.

@@

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
(Van De Schoot et al., 2017). While we concur that
investigating missingness is due diligence, it is important to emphasize
that missingness is adequately handled by default in \texttt{OpenMx}, which is
the computational backend of \texttt{tidySEM}. Its Full Information Maximum
Likelihood (FIML) estimator makes use of all available information
without imputing missing values, by estimating the model based on all
available information for each subject
(Enders, 2022; \textbf{endersRelativePerformanceFull2001?}). FIML is a
best-practice solution for handling missing data; at least on par with
multiple imputation (Lee \& Shi, 2021). Multiple imputation is less
suitable to LCA for two reasons. First, because LCA methods are often
computationally intensive, and conducting them on several imputed
datasets multiplies the load. Second, because many multiple imputation
methods assume normality; an assumption likely to be violated when data
come from a heterogeneous population.

Three types of missingness have been distinguished in the literature
(Enders, 2022; Rubin, 1976): random missingness
(MCAR); missingness contingent on \emph{observed} variables that can be known
and controlled for (MAR); and missingness related to unobserved
variables, which can be neither known nor controlled for (MNAR). A
so-called ``MCAR'' test is often reported (Jamshidian \& Jalal, 2010), but
note that this name is somewhat misleading, as a significant test
indicates that missingness is MAR, but a non-significant test statistic
does not distinguish between MCAR or MNAR. As the classic MCAR test
relies on the comparison of variances across groups with different
patterns of missing data, it assumes normality (Little, 1988). This
assumption is tenuous in the context of LCA. A non-parametric MCAR test
may be more suitable (Jamshidian \& Jalal, 2010). For this tutorial, the
lead author has implemented this test in the \texttt{mice} package as
\texttt{mice::mcar()}. Our recommendation is to inspect and report the
proportion of missingness per variable (e.g., using
\texttt{tidySEM::descriptives()}), and report the \texttt{mice::mcar()} test. As FIML
estimation assumes that missingness is either MCAR or MAR, one would
proceed with FIML regardless of the outcome.

@@

In principle, LCA can be based on any parametric model. At present,
\texttt{tidySEM} only facilitates estimation of models that can be specified in
\texttt{lavaan} syntax. The available model parameters depend on the
measurement level of the indicators. Continuous indicators can be
parametrized in terms of means, variances, and covariances. LCGA and GMM
have these same parameters, but with respect to the latent growth
variables. Ordinal indicators can be parametrized differently; the
default approach in \texttt{tidySEM} assumes that each ordinal indicator
reflects an underlying standard normal distribution,
\(N(\mu = 0, \sigma = 1)\). The parameters are ``thresholds'' that
correspond to quantiles of this distribution, estimated based on the
proportion of individuals in each of the response categories of the
indicator variable. For example, a binary indicator has a single
threshold that distinguishes the two response categories. If responses
are distributed 50/50, then the corresponding threshold would be
\(t_1 = 0.00\). If the responses are distributed 60/40, then the resulting
threshold would be \(t_1 = 0.25\). All of the aforementioned parameters
can be freely estimated, or constrained across classes, or fixed (e.g.,
to zero).

Prior literature emphasized the importance of considering alternative
model specifications
(\textbf{vandeschootGRoLTSChecklistGuidelinesReporting2017?}). What has remained
underemphasized is that alternative model specifications should be
approached differently for exploratory versus confirmatory analyses (see
Use Cases). Most prior literature has emphasized exploratory LCA (e.g., Nylund, Asparouhov, \& Muthén, 2007). In this context, model specification typically
consists of an exhaustive grid search along several model specifications
and varying numbers of classes. A ``final'' model is then selected based
on a combination of fit indices, significance tests, and
interpretability (see Class Enumeration).

Confirmatory LCA does not require such an extensive search over the
model space. If hypothesized models have been specified a priori, and
possibly even preregistered, it makes sense to focus on these models,
rather than proceed in a purely data-driven manner. Nonetheless, the
theoretical model specification should still be compared to a few
alternatives because LCA lacks objective fit indices. The fit of a
theoretical model can thus only be assessed by comparison to other
candidates. A one-class model is a sensible benchmark to test whether
the theoretical model fits better than a model that assumes a
homogeneous population. Other candidates for comparison include
competing theoretical models and those with a few more or less classes
than the hypothesized number. If the theoretical model has much better
fit than other models, for example, based on IC weights, this provides
evidence for the theory. If another model fits substantially better,
this provides evidence against the theory. If fit differences are small,
the theoretical model may still be preferred, as theory is a relevant
consideration in model selection (\textbf{jungIntroductionLatentClass2008?}). The
fact that the theoretical model received little differential support
should of course be discussed.

A second consideration in model specification is the risk of overfitting
(\textbf{hastieElementsStatisticalLearning2009?}). Overfitting occurs when a
model has many parameters relative to the number of observations,
causing it to capture idiosyncratic noise in the sample and limiting its
generalizability to new samples. As explained before, the number of
parameters in LCA scales with the number of estimated classes.
Consequently, LCA with many classes have a potentially very high number
of parameters. In Bayesian mixture modeling, it was observed that if the
estimated number of classes exceeded the true number, the number of
observations assigned to these classes tended towards zero as sample
size increased (\textbf{rousseauAsymptoticBehaviourPosterior2011?}). Although it
is not known whether frequentist LCA exhibits the same behavior, low
class counts are still an indicator of potential overfitting. The
section on Class Enumeration addresses ways to guard against
overfitting.

@@

One problem with the random starts approach is that it is
computationally expensive to run this many replications. Moreover,
because the algorithm begins with random starting values, it is
inefficient because many marbles are likely to start far away from any
valley. Additionally, if all sets of starting values are close to a
specific local optimum, there is a risk that the global optimum is never
found. An alternative solution that overcomes these challenges is to use
a global optimizer, like simulated annealing (SA, \textbf{coranaMinimizingMultimodalFunctions1987?}). SA does not behave quite like
the marbles of the random starts approach; instead, it iteratively
considers some ``destination'' in the landscape, and compares its \(-2LL\)
to the current one. If the destination \(-2LL\) is lower, the estimator
moves there. However, if the destination \(-2LL\) is \emph{higher}, it still
moves there occasionally, with a certain probability. Think of this as
occasionally flicking the marble, to see if it rolls back into the
current valley, or finds another, even deeper one. This property allows
SA it to escape local optima and find the global optimum.

@@

\hypertarget{follow-up-analyses}{%
\subsubsection{Follow-up Analyses}\label{follow-up-analyses}}
